---
title: "Finding differentially expressed genes using GSC cell lines"
author: "Sonali Arora"
date: "May 19, 2020"
output: 
  html_document:
    toc: true
    theme: united
---

In this vignette, we will compute calculate differentially expressed genes from the GSC cell lines. 

We first start with the G14 and G179 cell lines. We are looking for differentially expresssed
genes in G14 compared to G179 cell line samples.

# Compute Differentially expressed genes (DEGs)

```{r}

library(DESeq2)
library(ggplot2)
raw_counts_fname = "GSE89623_448_559_G14_G179.txt"
raw= read.delim(raw_counts_fname, header=T, stringsAsFactors = FALSE, 
                row.names=1, check.names = FALSE)

raw_counts = raw[, c("G14-1", "G14-2", "G14-3", 
              "G179-1", "G179-2", "G179-3")]

coldata = cbind(sampleName = colnames(raw_counts), 
                sampleGroup = c(rep("G14", 3), rep("G179", 3) ))

rownames(coldata) = colnames(raw_counts)
dds = DESeqDataSetFromMatrix(raw_counts, colData =coldata, 
                             design= ~sampleGroup)
dds$sampleGroup = relevel(dds$sampleGroup, "G179")
rld <- rlog(dds, blind=FALSE)

data <- plotPCA(rld,intgroup =c("sampleGroup"), ntop=nrow(raw_counts), returnData=TRUE)
percentVar <- round(100 * attr(data, "percentVar"))
p1=ggplot(data, aes(PC1, PC2, color=sampleGroup))+ geom_point(size=6) +
    xlab(paste0("PC1: ",percentVar[1],"% variance")) +
    ylab(paste0("PC2: ",percentVar[2],"% variance"))+
    geom_text(aes(label=sampleNames),hjust="inward", vjust=2, size=6)+
    scale_color_manual(breaks= c(control_group, test_group),
                       values=c("darkgreen", "purple")) 


dds = DESeq(dds)

res1 = results(dds, alpha=0.05, lfcThreshold = 1)
res1 = as.data.frame(res1)
res1 = res1[,c("log2FoldChange", "pvalue", "padj")]

G14_vs_G179_up_reg = res1[which(res1$log2FoldChange >= 1 & res1$padj < 0.05), ]
G14_vs_G179_down_reg = res1[which(res1$log2FoldChange <= -1 & res1$padj < 0.05), ]

```

Next, we calculate the number of differentially regulated genes in 448T compared to 559 cell line. 


````{r}

raw_counts = raw[, c( "448T-1", "448T-2", "448T-3",
             "559T-1" ,"559T-2" ,"559T-3")]

coldata = cbind(sampleName = colnames(raw_counts), 
                sampleGroup = c(rep("448", 3), rep("559", 3) ))

rownames(coldata) = colnames(raw_counts)
dds = DESeqDataSetFromMatrix(raw_counts, colData =coldata, 
                             design= ~sampleGroup)
dds$sampleGroup = relevel(dds$sampleGroup, "559")
rld <- rlog(dds, blind=FALSE)

data <- plotPCA(rld,intgroup =c("sampleGroup"), ntop=nrow(raw_counts), returnData=TRUE)
percentVar <- round(100 * attr(data, "percentVar"))
p2=ggplot(data, aes(PC1, PC2, color=sampleGroup))+ geom_point(size=6) +
    xlab(paste0("PC1: ",percentVar[1],"% variance")) +
    ylab(paste0("PC2: ",percentVar[2],"% variance"))+
    geom_text(aes(label=sampleNames),hjust="inward", vjust=2, size=6)+
    scale_color_manual(breaks= c(control_group, test_group),
                       values=c("darkgreen", "purple")) 

dds = DESeq(dds)

res2 = results(dds, alpha=0.05, lfcThreshold = 1)
res2 = as.data.frame(res2)
res2 = res2[,c("log2FoldChange", "pvalue", "padj")]

s448_vs_559_up_reg = res2[which(res2$log2FoldChange >= 1 & res2$padj < 0.05), ]
s448_vs_559_down_reg = res2[which(res2$log2FoldChange <= -1 & res2$padj < 0.05), ]

```


# GO analysis 

Now we will perform Gene Ontology (GO) and Pathway enrichment analysis (from Reactome)
on the differentially expressed genes. 

```{r}
source("Utils.R")

myEnrichmentFunction(rownames(s448_vs_559_up_reg), title="up_reg_in_448", resdir=getwd())
myEnrichmentFunction(rownames(s448_vs_559_down_reg), title="up_reg_in_448", resdir=getwd())

myEnrichmentFunction(rownames(G14_vs_G179_up_reg), title="up_reg_in_G14", resdir=getwd())
myEnrichmentFunction(rownames(G14_vs_G179_down_reg), title="up_reg_in_G14", resdir=getwd())

````


